---
title: 'Lab 09 - Importing Data'
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
library( dplyr )
library( pander )
```

#### [DS4PS](https://ds4ps.github.io/course_website/)

<br>

This lab introduces you to the basics of importing a dataset into R. You will be using the same dataset that we will use for the final project. You will also get a head start on the project by conducting some preliminary analysis of car crash data. 

You will need the following package for the lab:

```{r, eval=F}
library( dplyr )
```

You can create a new RMarkdown file, or download the lab template: 

[RMD Template](https://cdn.rawgit.com/DS4PS/Data-Science-Class/c464d527/TEMPLATES/MarkdownTemplateSimple.Rmd)

<br><br>

# Data

In this lab you will use a traffic accidents dataset from the Tempe Open Data Portal, available at:

https://data.tempe.gov/dataset/high-severity-traffic-crashes-1-08

You will download the "Crash Data" CSV dataset, store it on your machine, and read it into R using the read.csv() fuction.

The dataset contains the following variables:


|column              |type      |label                      |description                                                                                                                 |
|:-------------------|:---------|:--------------------------|:---------------------------------------------------------------------------------------------------------------------------|
|Incidentid          |numeric   |Incident ID                |Unique incident ID number assigned by Arizona Department of Transportation (ADOT).                                          |
|DateTime            |timestamp |Date Time                  |Date and time that the crash occurred.                                                                                      |
|Year                |numeric   |Year                       |Year that the crash occurred.                                                                                               |
|StreetName          |text      |Street Name                |The street that the crash occurred on.                                                                                      |
|CrossStreet         |text      |Cross-street               |The nearest intersecting street or road.                                                                                    |
|Distance            |numeric   |Distance from Intersection |The distance, in feet, that the crash occurred from the cross-street.                                                       |
|JunctionRelation    |text      |Junction Relation          |The location of the crash in relation to a junction, either an intersection or connection between a driveway and a roadway. |
|Totalinjuries       |numeric   |Total Injuries             |Total number of persons with non-fatal injuries involved in the crash.                                                      |
|Totalfatalities     |numeric   |Total Fatalities           |Total number of persons with fatal injuries involved in the crash.                                                          |
|Injuryseverity      |text      |Injury Severity            |The highest severity of injury of all persons involved in the crash.                                                        |
|Collisionmanner     |text      |Collision Manner           |Identifies the manner in which two vehicles initially came into contact.                                                    |
|Lightcondition      |text      |Lighting Conditions        |The type/level of light that existed at the time of the crash.                                                              |
|Weather             |text      |Weather                    |The prevailing (most significant) atmospheric conditions that existed at the time of the crash.                             |
|SurfaceCondition    |text      |Surface Condition          |The roadway surface condition at the time and place of a crash.                                                             |
|Unittype_One        |text      |Unit Type One              |Driver, Passenger, Pedestrian, Pedalcyclist or Driverless.                                                                  |
|Age_Drv1            |numeric   |                           |                                                                                                                            |
|Gender_Drv1         |text      |                           |                                                                                                                            |
|Traveldirection_One |text      |Travel Direction           |The direction the unit was traveling before the incident occurred,                                                          |
|Unitaction_One      |text      |Unit Action One            |The maneuver, or last action, of the unit before the crash.                                                                 |
|Violation1_Drv1     |text      |Violation  One             |The main violation/behavior of the unit that contributed to the crash.                                                      |
|AlcoholUse_Drv1     |text      |Alcohol Use 1              |Indicates whether alcohol was a contributing factor in the crash or not.                                                    |
|DrugUse_Drv1        |text      |Drug Use 1                 |Indicates whether drug use was a contributing factor in the crash or not.                                                   |
|Unittype_Two        |text      |Unit Type Two              |Driver, Passenger, Pedestrian, Pedalcyclist or Driverless.                                                                  |
|Age_Drv2            |numeric   |                           |                                                                                                                            |
|Gender_Drv2         |text      |                           |                                                                                                                            |
|Traveldirection_Two |text      |Travel Direction Two       |The direction the unit was traveling before the incident occurred.                                                          |
|Unitaction_Two      |text      |Unit Action Two            |The maneuver, or last action, of the unit before the crash.                                                                 |
|Violation1_Drv2     |text      |Violation Two              |The main violation/behavior of the unit that contributed to the crash.                                                      |
|AlcoholUse_Drv2     |text      |Alcohol Use 2              |Indicates whether alcohol was a contributing factor in the crash or not.                                                    |
|DrugUse_Drv2        |text      |Drug Use 2                 |Indicates whether drug use was a contributing factor in the crash or not.                                                   |
|Latitude            |numeric   |Latitude                   |Used to specify the precise location of the crash.                                                                          |
|Longitude           |numeric   |Longitude                  |Used to specify the precise location of the crash.                                                                          |





# Lab Instructions


Parts 1-4 are designed to give you practice acquiring data from an external site and reading it into R. Part 5 asks you to analyze the data to pick out some patterns.


**PART 1:** 

1. Download the dataset called "Crash Data" as a CSV file and save it to your computer. 
2. Write a code chunk to import the dataset into R using **setwd()** and **read.csv()**.
3. Show that you have correctly loaded the dataset by printing the variable names using the structure **str()** function. 

```{r, eval=F}
setwd( "C:/Users/username/Downloads" )
dat <- read.csv( "filename.csv" )
str( dat )
```

Note that if you copy the working directory address from a Windows folder it will use backslashes, whereas R follows the Unix convention of forward slashes.

Windows:

`C:\Users\usename\Downloads`

R:

`C:/Users/username/Downloads`

In R Studio you can set your working directory under Session --> Set Working Directory --> Choose Directory or using the shortcut Ctrl+Shift+H. Then get the correct R path by typing:

```{r, eval=F}
getwd()
```


**PART 2:** 

In the first step you are using what is called an "absolute path" reference. It is absolute because it gives the exact location (path) of the file on a specific machine. 

```{r, eval=F}
setwd( "C:/Users/JaneDoe/Downloads" )
```

Unfortunately, this causes problems if (1) you are collaborating with someone using a shared folder since their path will be different than your path, or (2) you are working on a project on both your laptop and your desktop and they have different usernames or directory structures for the project.  

As a result, it is ALWAYS recommended to use "relative path" references. When using R Studio, if you try to load a dataset it will look for the data in the same folder as the RMD file. So one option is to save the dataset into your labs folder, then read it directly from there without setting the working directory:

```{r, eval=F}
dat <- read.csv( "filename.csv" )
str( dat )
```

If you have lots of raw datasets (which is common in big projects) your folders quickly get cluttered. It is good to get into the habit of saving all of your data in a data folder to keep your project organized. You can always reference files inside of a folder using a partial path as follows:

```
CPP-591
  ├── Labs
  |   ├──Lab-09.Rmd
  |   ├── data
  |   |   ├──filename.csv
```

```{r, eval=F}
dat <- read.csv( "data/filename.csv" )
```

Or else:

```{r, eval=F}
setwd( "data" )
dat <- read.csv( "filename.csv" )
```

Conversely, if your data folder is not inside of the folder, you can navigate up one level using ..


```
CPP-591
  ├── Labs
  |   ├──Lab-09.RMD
  ├── data
  |   ├──filename.csv
```

```{r, eval=F}
dat <- read.csv( "../data/filename.csv" )
```


Create a folder called "data" in your lab folder, and repeat PART 1 above WITHOUT the **setwd()** command by using a relative path. Read in the data file and print out variables using **str()**.


**PART 3:**

Note that the open data site provides a URL for the dataset. If you want your code to be as general as possible, you can load the data directly from the web link instead of from a local file, and use **str()** to print out variable names.

```{r, eval=F}
URL <- "dataset_web_address_here"
dat <- read.csv( URL )
str( dat )
```


**PART 4:**

Note the variable classes in the previous steps. By default R tries to guess which variables are factors, and it doesn't often do a good job. Load the data again, but this time add the argument stringsAsFactors=FALSE to the **read.csv()** function, and re-print dataset names and classes using **str()**. What has changed? 



**PART 5:**

You are an analyst employed by the City of Tempe. The Mayor's office sends a request for some information about the causes of injuries and fatalities in traffic accidents in the city. Use the Crash data and dplyr functions to answer the following questions:

1. How many injuries and fatalities occured in 2017? 
2. Historically, what proportion of car accidents in Tempe result in injuries or fatalities? 
3. What sort of accident ("Collisionmanner") is most likely to result in an injury or fatality? 
4. Of all of the accidents that result in injuries or fatalities, how often are alcohol and drugs involved?
5. Do young drivers (under the age of 26) or elderly drivers (over the age of 64) cause more accidents? Assume the first drive caused the accident.  
6. Are rates of injury or fatality different for those two groups? In other words, what proportion of accidents caused by each group result in injuries or fatalities? 
7. What are the three most common types of accidents caused by each group? 










# Submission Instructions

After you have completed your lab, knit your RMD file.


Login to Canvas at <http://canvas.asu.edu> and navigate to the assignments tab in the course repository. Upload your RMD and your HTML files to the appropriate lab submission link.

Remember to:

* name your files according to the convention: **Lab-##-LastName.Rmd**
* show your solution, include your code.
* do not print excessive output (like a full data set).
* follow appropriate style guidelines (spaces between arguments, etc.).
