---
title: 'Lab 02 - Functions and Vectors'
output:
  html_document:
    theme: united
    df_print: paged
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
library( dplyr )
library( pander )
```

#### [DS4PS](https://ds4ps.github.io/course_website/)

<br>

*This lab offers practice with some basic R functions applied to vectors in order to ascertain basic information about the dataset or count and summarize things.*


<br><br>

## Functions

You will use the following functions for this lab: 

```{r,eval=F}
names()                 # variable names
head()                  # preview dataset
length()                # vector dimensions
dim(), nrow(), ncol()   # dataset dimensions
sum(), summary()        # summarize numeric vectors
table()                 # summarize factors / character vectors
```

Recall that referencing variables in R requires the dataset name and variable name, along with the `$` operator.

```{r, eval=F}
summarize( dat$acres )
```



<br><br>

# Data

This lab will use [city tax parcel data](https://ds4ps.github.io/Data-Science-Class/DATA/syr_parcels.html) from Syracuse, NY.

```{r}
URL <- "https://raw.githubusercontent.com/DS4PS/Data-Science-Class/master/DATA/syr_parcels.csv"
dat <- read.csv( URL, stringsAsFactors=FALSE )
```

```{r, echo=F}
head( dat, 10 ) 
```

<br><br>

# Lab Instructions

Answer the following questions in your lab.  

1. How many tax parcels are in syracuse (dataset dimensions)? 
2. How many acres of land are in syracuse (sum over a numeric vector)?
3. How many vacant BUILDINGS are there in Syracuse (sum over a logical vector)?
4. What proportion of parcels are tax exempt?
5. Which neighborhood contains the most tax parcels (table with one variable)?
6. Which neighborhood contains the most vacant LOTS (table with two variables)? 

*Recall that R wants to make sure you are aware of missing values, so it will return NA for functions performed on vectors with missing values. Add the 'NA remove' argument to these vectors to ignore missing values:* `sum( x, na.rm=T )`



<br><br>

# Submission Instructions

You can create a new RMarkdown file, or download a template. 

[RMD Template](https://cdn.rawgit.com/DS4PS/Data-Science-Class/d7a11a0e/TEMPLATES/MarkdownTemplate.RMD)

When you have completed your assignment, knit your RMD file to generate your rendered HTML file. Platforms like BlackBoard and Canvas often disallow you from submitting HTML files when there is embedded computer code, so create a zipped folder with both the RMD and HTML files.

Login to Canvas at <www.canvas.asu.edu> and navigate to the assignments tab in the course repository. Upload your zipped folder to the appropriate lab submission link.

Remember to:

* name your files according to the convention: **Lab-##-LastName.Rmd**
* show your solution, include your code.
* do not print excessive output (like a full data set).
* follow appropriate style guidelines (spaces between arguments, etc.).

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples.

Note that when you knit a file, it starts from a blank slate. You might have packages loaded or datasets active on your local machine, so you can run code chunks fine. But when you knit you might get errors that functions cannot be located or datasets don't exist. Be sure that you have included chunks to load these in your RMD file.

Your RMD file will not knit if you have errors in your code. If you get stuck on a question, just add `eval=F` to the code chunk and it will be ignored when you knit your file. That way I can give you credit for attempting the question and provide guidance on fixing the problem.


