---
title: 'Lab 03 - Logical Statements'
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
library( dplyr )
library( pander )
```

#### [DS4PS](https://ds4ps.github.io/course_website/)

<br>

*This lab offers practice using logical operators to translate management questions from English into computer code.*

You can create a new RMarkdown file for this lab, or download the lab template: 

[RMD Template](https://cdn.rawgit.com/DS4PS/Data-Science-Class/a48d67a5/TEMPLATES/MarkdownTemplate.RMD)



<br><br>

## Functions

You will use the following operators and functions for this lab: 

```{r,eval=F}
logical operators  ( ==, >, <, etc. )
AND and OR statements
subset operator [ ] 
sum()        # count TRUEs in logical vectors
mean()       # proportion of TRUEs in logical vectors
```

Recall that referencing variables in R requires the dataset name and variable name, along with the `$` operator.

```{r, eval=F}
summary( dat$acres )
```



<br><br>

# Data

This lab will use [city tax parcel data](https://ds4ps.github.io/Data-Science-Class/DATA/syr_parcels.html) from Syracuse, NY.

```{r load_data}
URL <- "https://raw.githubusercontent.com/DS4PS/Data-Science-Class/master/DATA/syr_parcels.csv"
dat <- read.csv( URL, stringsAsFactors=FALSE )
```

```{r, echo=F}
head( dat, 10 ) 
```

<br><br>


## Data Dictionary

VARIABLE | DESCRIPTION
-------- | ----------------------
**tax_id**  |  unique parcel ID  (numeric)  
**neighborhood**  |  city neighborhood for tax parcel (character) 
**stnum**  |  parcel address (numeric) 
**stname**  |  parcel address (character) 
**zip**  |  parcel address (numeric) 
**owner**  |  property owner (character) 
**frontfeet**  |  length of the front of the lot (numeric)    
**depth**  |  depth of the lot (numeric)
**sqft**  |  size of the parcel in square feet  (numeric)
**acres**  |  size of the parcel in acres (numeric)
**yearbuilt**  |  year of construction for the building on the parcel (numeric) 
**age**  |  age of the building (numeric)
**age_range**  |  age of the building by decade (factor)  
**land_use**  |  zoning purpose of the parcel (factor) 
**units**  |  number of units in multi-unit buildings (numeric)        
**residential**  |  is the parcel zoned for residential (T/F)  
**rental**  |  is the residential space zoned for rental (T/F) 
**vacantbuil**  |  is the building vacant (T/F)   
**assessedla**  |  assessed value of the parcel land  (numeric)
**assessedva**  |  assessed value of the parcel land plus buildings (numeric) 
**tax.exempt**  |  is the parcel tax exempt (T/F)   
**countytxbl**  |   value of parcel for tax purposes (numeric)  
**schooltxbl**  |  value of parcel for tax purposes (numeric) 
**citytaxabl**  |  value of parcel for tax purposes (numeric)   
**star**  |  does the parcel qualify for a residential tax break (T/F)      
**amtdelinqu**  |  amount of tax bill that is delinquint (numeric) 
**taxyrsdeli**  |  length of time tax bill is delinquint (numeric) 
**totint**  |  total interest owed for delinquint taxes (numeric) 
**overduewater**  |  total owed on overdue water bills (numeric) 


<br><br>


# Lab Instructions

Answer the following questions in your lab by writing the appropriate logical statements. Practice using compound statements and subset operators.  

1. How many tax parcels in Syracuse are tax-exempt? 
3. How many acres of land are in Syracuse? 
3. What proportion of total land in Syacuse is tax exempt (use `acres`)?

4. How many single family homes were built between 1920 and 1945? 
5. How many houses in the Westcott neighborhood have an assessed value (assessedva) of over $200,000? 

5. What proportion of tax bills are delinquent? 
6. What is the average amount owed by those with delinquint tax bills? 

7. What proportion of tax parcels have an overdue water bill (overduewater)?
8. Which type of land use is category has the most overdue water bills (selector vector plus table)? 


*Recall that R wants to make sure you are aware of missing values, so it will return NA for functions performed on vectors with missing values. Add the 'NA remove' argument to these vectors to ignore missing values:* `sum( x, na.rm=T )`



<br><br>

# Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file. Platforms like BlackBoard and Canvas often disallow you from submitting HTML files when there is embedded computer code, so create a zipped folder with both the RMD and HTML files.

Login to Canvas at <http://canvas.asu.edu> and navigate to the assignments tab in the course repository. Upload your zipped folder to the appropriate lab submission link.

Remember to:

* name your files according to the convention: **Lab-##-LastName.Rmd**
* show your solution, include your code.
* do not print excessive output (like a full data set).
* follow appropriate style guidelines (spaces between arguments, etc.).

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples.

## Notes on Knitting

Note that when you knit a file, it starts from a blank slate. You might have packages loaded or datasets active on your local machine, so you can run code chunks fine. But when you knit you might get errors that functions cannot be located or datasets don't exist. Be sure that you have included chunks to load these in your RMD file.

Your RMD file will not knit if you have errors in your code. If you get stuck on a question, just add `eval=F` to the code chunk and it will be ignored when you knit your file. That way I can give you credit for attempting the question and provide guidance on fixing the problem.