---
title: 'Lab 05 - Intro to plot()'
output:
  html_document:
    theme: readable
    df_print: paged
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=F, warning=F)
library( dplyr )
library( pander )
```

#### [DS4PS](https://ds4ps.github.io/course_website/)

<br>

*You lab introduces students to the plot() function.*

You can create a new RMarkdown file, or download the lab template: 

[RMD Template](https://cdn.rawgit.com/DS4PS/Data-Science-Class/a48d67a5/TEMPLATES/MarkdownTemplate.RMD)



<br><br>

## Functions

You will use the **plot()** function for this lab and have a chance to practice common arguments: 

```{r, eval=F}
plot(
  
  x=fertilizer,          # data, if x is omitted then uses 1:length(y)
  y=corn.height,                   
  xlim=c(0,100),         # min and max value of axes
  ylim=c(200,350),   
  frame.plot=FALSE,      # draw a box around the data?
  col="steel blue",      # color of the points
  type="b",              # points="p", lines="l", both="b", or none="n"
  pch=19,                # shape of points to plot
  cex=2,                 # size of points
  main="Plot Title",     # title of your plot
  xlab="Label for X",    # axes labels
  ylab="Label for Y",  
  cex.lab=1.5            # aspect ratio for axes labels
  
  )
```




<br><br>

# Data

This data comes from the **AEA** package that accompanies the textbook:

*Christian Kleiber and Achim Zeileis (2008), Applied Econometrics with R, Springer-Verlag, New York.*

We are using the **CASchools** dataset:

*The data used here are from all 420 K-6 and K-8 districts in California with data available for 1998 and 1999. Test scores are on the Stanford 9 standardized test administered to 5th grade students. School characteristics (averaged across the district) include enrollment, number of teachers (measured as “full-time equivalents”, number of computers per classroom, and expenditures per student. Demographic variables for the students are averaged across the district.*

We will use the following four variables, copied to GitHub for convenience:

VARIABLE | DESCRIPTION
-------- | ----------------------
**test**  |  Average standardized math and reading test scores for students at each school 
**free.lunch** | Proportion of students at each school that qualify for free or reduced lunches
**student.ratio** | The student-to-teacher ratio of each school
**per.student.exp** | The total school budget divided by the number of students 


```{r load_data}
URL <- "https://raw.githubusercontent.com/DS4PS/Data-Science-Class/master/DATA/CASchools.R"
source( URL )
```


```{r, echo=F}
library( pander )
model.dat <- data.frame( test, student.ratio, per.student.exp, free.lunch) 

desc.stats <- data.frame( N = sapply( model.dat, function(x){sum(!is.na(x))} ),
                          Min = sapply( model.dat, min ),
                          Q25 = sapply( model.dat, quantile, 0.25, na.rm=T ),
                          Mean = sapply( model.dat, mean ),
                          Q75 = sapply( model.dat, quantile, 0.75, na.rm=T ),
                          Max = sapply( model.dat, max ) )

desc.stats <- round( desc.stats, 2 )

rownames(desc.stats) <- c("Student Combined Test Scores (Math and Reading)",
                              "Student Teacher Ratios",
                              "Ave. School Expenditure Per Student",
                              "Percentage of Students Eligible for Free Lunches")

desc.stats %>% pander()
```




<br><br>








# Lab Instructions

We are interested in predictors of school test scores. 

We have three independent variables for our study: school budgets on a per-student basis, student teacher ratios, and proportion of students in the school that qualify for free or reduced lunches (a proxy for poverty status). 

1. Create three scatterplots to visualizing the relationship between our three study variables and test scores. Test scores should be on the y-axis of each.

2. Select a different style (pch=) and color (col=) for plot points on each graph.

3. Add meaningful variable labels and plot titles to each graph.

4. Which of these variables is the best predictor of test scores? Explain your answer. 

*You are free to adjust the rest of your graph aestetics as desired.* 


<br><br>

# Submission Instructions

When you have completed your assignment, knit your RMD file to generate your rendered HTML file. Platforms like BlackBoard and Canvas often disallow you from submitting HTML files when there is embedded computer code, so create a zipped folder with both the RMD and HTML files.

Login to Canvas at <http://canvas.asu.edu> and navigate to the assignments tab in the course repository. Upload your zipped folder to the appropriate lab submission link.

Remember to:

* name your files according to the convention: **Lab-##-LastName.Rmd**
* show your solution, include your code.
* do not print excessive output (like a full data set).
* follow appropriate style guidelines (spaces between arguments, etc.).

See [Google's R Style Guide](https://google.github.io/styleguide/Rguide.xml) for examples.

## Notes on Knitting

Note that when you knit a file, it starts from a blank slate. You might have packages loaded or datasets active on your local machine, so you can run code chunks fine. But when you knit you might get errors that functions cannot be located or datasets don't exist. Be sure that you have included chunks to load these in your RMD file.

Your RMD file will not knit if you have errors in your code. If you get stuck on a question, just add `eval=F` to the code chunk and it will be ignored when you knit your file. That way I can give you credit for attempting the question and provide guidance on fixing the problem.