# MAPS {-}

This chapter introduces the use of map tiles to provide context for analysis of geographic point data in R, data usually represented by Latitude and Longitude measures on the X-Y coordinate system:

```{r, echo=F}
library( ggmap )
library( dplyr )

dat <- readRDS( "../DATA/TempeTrafficAccidents.rds" )
d2 <- 
  dat %>% 
  filter( Totalinjuries > 0 | Totalfatalities > 0 ) %>%
  mutate( tot.injuries.fatalities = Totalinjuries + Totalfatalities )

# summary( d2$tot.injuries.fatalities )

par( mar=c(0,0,4,0) )
plot( d2$Latitude, d2$Longitude, pch=19, 
      cex=0.25*d2$tot.injuries.fatalities, 
      col=alpha( "firebrick", alpha=0.5),
      main="Traffic Accidents in Tempe, AZ",
      xlab="", ylab="",
      bty="n", axes=F )

title( main="(plot size relative to injuries and fatalities)", 
       line=-0.3, cex.main=0.7 )

map <- get_map( location="tempe az" )
ggmap( map )
```





# Map Tiles

We will be using the **ggmap** package to  

```
install.packages( "ggmap" )
```

The packapge provides a tremendous amount of GIS and mapping functionality, but for now we will utilize a built-in API that allows us to access map tiles from a variety of commercial and open-source databases.

HTML maps are comprised of individual square image "tiles" stored in a database that can query tiles by their geographic coordinates. When the user requests a specific map, the appropriate tiles are downloaded and assembled to create the map. Unique databases exist for each type of map (terrain vs satellite, for example), and each aerial vantage point (distance from the earth).

Map tiles are background images only. They cannot be used to create choropleth maps or to conduct analysis. They provide context only for the data that is layered over the top of the images. 

In order to build an HTML map, the database query requires the one of the following:

* A bounding box which specifies the four corners of the map.

OR

* One point on a map and a "zoom" level.

AND 

* The type of map aesthetic.


The functions we will use today are 'get_map' and the function 'ggmap' (which is a little confusing because it is both the function and the package name, but follow along below and it will be clear).

Let's see how it will work.

```{r}

library( ggmap )

map <- get_map( location="syracuse ny" )

ggmap( map )

```



Excellent, with two lines of code we have created a map of our city. This is an extremely powerful function as it utilizes APIs from four major map archives, and automatically syncs the map images to a lattitude and longitude x-y axis. This will be important when we start to geocode data (covert addresses into lat-long coordinates to place points on maps).

The basemaps that you are generating here are simple image tiles. They are not GIS maps, which consist of a special format called shapefiles. Base maps provide a geographic context, but you cannot edit elments on the image directly.

# Arguments of get_map

Let's take a look at the core arguments of the function. If you look at the documentation you will see that there are 11 separate arguments necessary to execute the function. Fortunately, all have reasonable defaults so the function is pretty user friendly. 



```

get_map( 
         location = c(lon = -95.3632715, lat = 29.7632836),
         source = c("google", "osm", "stamen", "cloudmade")         
         zoom = "auto", 
         maptype = c("terrain", "satellite", "roadmap", "hybrid", "toner", "watercolor"), 
         color = c("color", "bw")
       )
```



Lets look at common usage of the arguments.

# Setting the location

The get_map function is clever in that it understands two formats of the location argument. You can either state a common name of a specific location. Or you can use a precise latitude and longitude coordinate.

When you use the name of your location, such as "Syracuse ny", the function links to a google maps API and searches for your term. Note that you need to be careful since your search may not yield the expected result. For example, look at what happens when I search for the Maxwell School.

```{r}
map <- get_map( location="maxwell school" )
ggmap( map )
```

Hmmm.... that is odd. That is not a map of Syracuse NY. What's going on? Well, if you search google maps you will see that Maxwell Elementary School in Greensburg, PA is one of the first results. 

You might try being a little more specific:

```{r}
map <- get_map( location="maxwell school of citizenship" )
ggmap( map )
```

That looks better.

Let's try the other format for the 'location' argument. If I look up Syracuse University in a GIS database I can determine that the latitude and longitude coordinates are: lat = 43.038161, lon = -76.135584. Let's try an explicit call to the map function:

```{r}
map <- get_map( location=c(lon = -76.135584, lat = 43.038161) )
ggmap( map )
```

We can see that it works as expected.






# Selecting a map source

We have several different map sources to chose from:

[Google Maps](https://developers.google.com/maps/)

[Open Streem Maps](http://en.wikipedia.org/wiki/OpenStreetMap)

[CloudMadeMaps](http://en.wikipedia.org/wiki/CloudMade)

[Stamen Maps](http://stamen.com/clients/cloudmade)

Unfortunately, CloudMade and Stamen have recently become proprietary map sources, so you need a subscription in order to access their maps. We limit our map-making to the Google and Open Street Map archives.

You change the map archive using the 'source' argument:


```{r}
map <- get_map( location="syracuse university", source="osm" )

ggmap( map )
```

We can see that the style of map has changed from the familiar default google maps tile.

Note, you may get the following error when using OSM maps:

> cannot open: HTTP status was '503 Service Unavailable'

This means that too many people are trying to access maps via the OSM servers (or they are down for another reason) and you need to try your request later. The API for OSM can be hit and miss.




# Map scale - the 'zoom' argument


The 'zoom' argument sets the scale of the map. From the documentation of the get_map function we see that:

> map zoom, an integer from 3 (continent) to 21 (building), default value 10 (city). openstreetmaps limits a zoom of 18.

Let's look at how we might create a map of Syracuse University.



```{r}
map <- get_map( location="syracuse university",
                source="google",
                zoom=21 )
ggmap( map )
```

Well, that's not very informative. Let's zoom out a little bit.

```{r}
map <- get_map( location="syracuse university",
                source="google",
                zoom=18 )
ggmap( map )
```

That's better. What about in Open Street Map format?

```{r}
map <- get_map( location="syracuse university",
                source="osm",
                zoom=18 )
ggmap( map )
```





# Changing the style of the map with the 'maptype' argument

We might think about changing the style of the base map to offer different context for the analysis. If we are presenting an urban planning topic the street grid is likely important. If we are presenting environmental analysis the satellite image of the terrain might be preferable. We have several styles to chose from:

> maptype = c("terrain", "satellite", "roadmap", "hybrid", "toner", "watercolor")

Note within the documentation that certain styles can only be used with certain base map sources. Google provides four types of maps (terrain, satellite, roadmap and hybrid), whereas Stamen Maps provides three (terrain, watercolor, toner). OSM has only one template.

> character string providing map theme. options available are 'terrain', 'satellite', 'roadmap', and 'hybrid' (google maps), 'terrain', 'watercolor', and 'toner' (stamen maps), or a positive integer for cloudmade maps (see ?get_cloudmademap)

```{r}

ggmap( get_map( location="baltimore", source="google", maptype="terrain", zoom=13 ) )

ggmap( get_map( location="baltimore", source="google", maptype="satellite", zoom=13 ) )

ggmap( get_map( location="baltimore", source="google", maptype="hybrid", zoom=13 ) )


```



# Creating black and white maps


```{r}

ggmap( get_map( location="baltimore", source="google", maptype="terrain", zoom=13, color="bw" ) )

ggmap( get_map( location="baltimore", source="google", maptype="satellite", zoom=13, color="bw" ) )

ggmap( get_map( location="baltimore", source="google", maptype="hybrid", zoom=13, color="bw" ) )


```





# Adding Points to a Map

So far you have learned how to download a base map from one of the various public map archives like Google maps or OpenStreetMap. The base map serves as context for our analysis. We now need to add datat to the map.

GIS maps present data in two ways. Individual observations can be plotted as points on the map. This includes things like location data such as a museum or the location of a robbery, or it can include things like physical addresses of households. Points are added to a map using the same convention as a scatterplot where longitudinal coordinates comprise the x-axis and latitude coordinates comprise the y-axis.

Alternatively, data can be aggregated and summarized by levels per geographic units. We might report average income for each census tract, number of crimes per county, or high school graduation rates per state. It could also be binary, such as whether a state voted republican or democrat in a specific election. In each case the data needs to be aggregated by a specific geographic unit, then the geography is shaded to represent a specific level. These are called choropleth maps. They will be covered in Lab 07.


We will use the *geom_point* function in the *ggmap* package to add points to our plot. It requires the following arguments:

Argument | Values
-------- | -------
data=    | Name of the data set that contains the latitude and longitude coordinates.
aes=     | The name of your latitude and longitude variables in the data set.
size=    | The size of your points with 1 being default, 2 being double that, etc.
col=     | The color of your points. For options type *colors()*.
alpha=   | THe level of transparency of your points. Accepts values between 0 (invisible) and 1 (solid).

```{r}

library( ggmap )

lat <-  43.03915
lon <- -76.13512

lat.long <- data.frame( lat, lon )

# lat.long <- geocode( 'syracuse university' ) creates the same results, see below

ggmap( get_map(  location='syracuse university', zoom = 12, color="bw" ), extent="device" ) +

geom_point( data=lat.long, aes(x=lon, y=lat), size=2, col="red", alpha=1 )



ggmap( get_map(  location='syracuse university', zoom = 12, color="bw" ), extent="device" ) +

geom_point( data=lat.long, aes(x=lon, y=lat), size=50, col="blue", alpha=0.3 )



```






# Data that includes latitude and longitude coordinates

Some data sets come with pre-set location coordinates. These data sets are easy to work with because all we need to do is use the correct variable names in our *geom_point* function and we are good to go.

One example is the New York City database of 311 service calls:

https://data.cityofnewyork.us/Social-Services/311-Service-Requests-2009/3rfa-3xsf

Note that this is a large dataset (1,783,133 observations). It will take awhile if you try to download it!


# Geocoding address data

Most datasets, however, do not come with pre-set location coordinates. The most typical kind of location data is a mailing address or a street intersection. If we want to add these data as points on a map we must first translate the address information into a coordinate system. This is called geocoding.

For this example let's use the list of farmers markets active in NYC in 2012:

https://data.cityofnewyork.us/Business/2012-NYC-Farmers-Market-List/b7kx-qikm

We will read this data into R, process the address information, and geocode it to create latitude and longitude coordinates.

First, let's see how the *paste* function works. We will use it to add city and state fields to the address data.

```{r}
# We use the paste function to combine separate text fields into a single string.

paste( "Marlon", "Brando", sep=" " )

paste( "Brando", "Marlon", sep=", " )

paste( "houses", "Price", sep="$" )

# It also works with columns of data

x <- 1:3

y <- c("a","b","c")

paste( x, y, sep="" )

```

Ok, now let's see a geocoding example.

```{r}

# install RCurl if you have not done this already using install.packages("RCurl")

library( RCurl ) # package to read data from html files

# copy the download link by right-clicking on the 'download as csv' link

# remove the "s" in "https"

url1 <- "http://data.cityofnewyork.us/api/views/b7kx-qikm/rows.csv?accessType=DOWNLOAD"

farmers.markets <- read.csv( file=url1, stringsAsFactors=F )

names( farmers.markets )

dim( farmers.markets ) # 137 markets in our data set

# We will use the 'Street.Address' field to create a full address

head( farmers.markets$Street.Address )

addresses <- paste( farmers.markets$Street.Address, "New York", "NY", sep=", " )

head( addresses )


```

Now that we have a vector of addresses, let's geocode them using the *ggmap* package.

```{r}

# I am suppressing messages here so the document is not full of non-informative output from Google.
#
# To see warnings you would simply run:
#
# market.coords <- geocode( addresses )

options( warn=-1 )

suppressMessages( market.coords <- geocode( addresses, messaging=F ) )
  
ggmap( get_map("Queens, NY", col="bw", zoom=11 ), extent="device" ) +

  geom_point( data=market.coords, aes(x=lon, y=lat), size=5, col="red", alpha=0.5 )

# To add the location data back to your original dataset use cbind()

farmers.markets <- cbind(farmers.markets, market.coords)

```

Note that the Google API only allows you to map 2,500 addresses each day, so there are limits to how much data you can geocode at any one time. 

If you were doing a larger project (many thousands of observations) you would have them geocoded through a commercial application, add the location data to your dataset, and then plot the points.



# Heat Maps

Heat maps are a nice way to summarize point data by highlight areas with dense clusters of points.

For this example we will use public art projects funded by NYC:

https://data.cityofnewyork.us/Recreation/Completed-Percent-for-Art-projects-with-artist-inf/gzdv-qiga

We will create a heat map of projects to show areas with the highest concentation of public art, then add yellow points for the actual projects over the top. 


```{r}

url2 = "http://data.cityofnewyork.us/api/views/gzdv-qiga/rows.csv?accessType=DOWNLOAD"

my.data = read.csv(file=url2, stringsAsFactors=F )

library( ggmap )

full.address <- paste( my.data$Address, my.data$City, "NY", sep=", " )

head( full.address )

suppressMessages( location <- geocode( full.address ) )

nyc <- ggmap( get_map( "Queens, NY", color="bw", zoom=11 ), extent="device" )

h2 <- nyc + stat_density2d( data=location, aes(x=lon,y=lat, fill=..level..), bins=50, geom="polygon", alpha=0.1 )

h3 <- h2 + geom_point(data=location, aes(x=lon,y=lat),size = 2, alpha=0.3, col="goldenrod" )

h3


```

